### 2024-01-04 sections/1-introduction.tex
tweaks from RFD

Examples of large datasets include neural network embeddings, genomic databases and time-series data such as radio frequency signals.
Large language models such as GPT~\cite{2020arXiv200514165B, OpenAI2023GPT4TR} and LLAMA-2~\cite{Touvron2023Llama2O}, and image embedding models~\cite{radford2021learning, dosovitskiy2020image} are a common source of neural network embeddings.
These embeddings are often high-dimensional, and the sizes of training and inference datasets for such networks are growing exponentially.
Among Biological datasets, the GreenGenes project~\cite{desantis2006greengenes} provides a multiple-sequence alignment of over one million bacterial 16S sequences, each 7,682 characters in length while Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 million genomes with an aligned length of 50,000 letters.
Among time-series datasets, the RadioML dataset~\cite{oshea2018radioml} contains approximately 2.55 million samples of synthetically generated signal captures of different modulation modes over a range of SNR levels.

@@

For example, for a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but may be sensitive to local perturbations for smaller values of $k$.
This is especially true when classes are not well-separated~\cite{zhang2022imbalanced}.
Further, there is some evidence that distance functions which do not obey the triangle inequality, such as cosine distance, perform poorly for $k$-NN search in biomedical settings~\cite{hu2016distance};
this suggests that approximate $k$-NN search could exhibit suboptimal classification accuracy in such contexts.

$\rho$-NN search also has a variety of applications.
For example, one could search for all genomes within a maximum edit distance of a query genome to find evolutionarily-related organisms~\cite{budowski2010fragbag}.
One could also search for all words within a maximum edit distance of a misspelled word to suggest corrections~\cite{ukkonen1985algorithms}.
Given an advertisement and a database of user profiles, one could search for all users whose profiles are ``similar enough'' to serve the advertisement to~\cite{zhang2020privacy}.
GPS and other location-based services use $\rho$-NN search to find nearby points of interest~\cite{zhang2020privacy} to a user's location.
In all of these cases, approximate $\rho$-NN search may be sufficient, but exact $\rho$-NN search is preferable.

This paper focuses on three novel algorithms for exact $k$-NN search.
We also present some improvements to the clustering and $\rho$-NN search algorithms in CHESS~\cite{ishaq2019clustered}, as well as improved genericity across distance functions.
We provide a comparison of CAKES to several state-of-the-art algorithms for similarity search, FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on several datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We also benchmark CAKES on a genomic dataset, the Silva 18S dataset~\cite{10.1093/nar/gks1219} using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and a radio frequency dataset, the RadioML dataset~\cite{oshea2018radioml} using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.

@@

Recent search algorithms for tackling the exponential growth of data include Hierarchical Navigable Small World networks (HNSW)~\cite{Malkov2016EfficientAR}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, random projection and tree building (ANNOY)~\cite{annoy}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}.

>>>>

Examples of large datasets include genomic databases, time-series data such as radio frequency signals, and neural network embeddings.
Large language models such as GPT~\cite{2020arXiv200514165B, OpenAI2023GPT4TR} and LLAMA-2~\cite{Touvron2023Llama2O}, and image embedding models~\cite{radford2021learning, dosovitskiy2020image} are a common source of neural network embeddings.
These embeddings are often high-dimensional, and the sizes of training and inference datasets for such networks are growing exponentially.
Among biological datasets, the GreenGenes project~\cite{desantis2006greengenes} provides a multiple-sequence alignment of over one million bacterial 16S sequences, each 7,682 characters in length while SILVA 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 million genomes with an aligned length of 50,000 letters.
Among time-series datasets, the RadioML dataset~\cite{oshea2018radioml} contains approximately 2.55 million samples of synthetically generated signal captures of different modulation modes over a range of SNR levels.

@@

For example, for a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but may be sensitive to local perturbations for smaller values of $k$.
This is especially true when classes are not well-separated~\cite{zhang2022imbalanced}.
Further, there is evidence that distance functions which do not obey the triangle inequality, such as cosine distance, perform poorly for $k$-NN search in biomedical settings~\cite{hu2016distance};
this suggests that approximate $k$-NN search could exhibit suboptimal classification accuracy in such contexts.

$\rho$-NN search also has a variety of applications.
For example, one could search for all genomes within a maximum edit distance of a query genome to find evolutionarily-related organisms~\cite{budowski2010fragbag}.
One could also search for all words within a maximum edit distance of a misspelled word to suggest corrections~\cite{ukkonen1985algorithms}.
Given an advertisement and a database of user profiles, one could search for all users whose profiles are ``similar enough'' to target the advertisement~\cite{zhang2020privacy}.
GPS and other location-based services use $\rho$-NN search to find nearby points of interest~\cite{zhang2020privacy} to a user's location.
In all of these cases, approximate $\rho$-NN search may be sufficient, but exact $\rho$-NN search is preferable.

This paper focuses on three novel algorithms for exact $k$-NN search.
We also present some improvements to the clustering and $\rho$-NN search algorithms in CHESS~\cite{ishaq2019clustered}, as well as improved genericity across distance functions.
We provide a comparison of CAKES to several state-of-the-art algorithms for similarity search, FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We also benchmark CAKES on a genomic dataset, the SILVA 18S dataset~\cite{10.1093/nar/gks1219} using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and a radio frequency dataset, the RadioML dataset~\cite{oshea2018radioml} using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.

@@

Recent search algorithms designed to scale with the exponential growth of data include Hierarchical Navigable Small World networks (HNSW)~\cite{Malkov2016EfficientAR}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, random projection and tree building (ANNOY)~\cite{annoy}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. However, some of these algorithms do not provide exact search.

### 2024-01-04 sections/2-methods.tex
tweaks from RFD

We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the Silva 18S rRNA dataset which contains which contains genomes with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, due to physical constraints (namely, biological evolution and the chemistry of RNA replication and mutation processes), the data are constrained to a lower-dimensional manifold within this embedding space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets obey the manifold hypothesis.

@@

It remains to provide an estimate for $\rho_k$. 
For this, we once again rely on the assumption that the query is drawn from the same distribution as the rest of the data, and thus the LFD at the query point is not significantly different from the LFD of nearby clusters.

@@

By Equation~\ref{eq:methods:lfd-original}, we then have that

>>>>

We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset which contains which contains genomes with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, due to physical constraints (namely, biological evolution and the chemistry of RNA replication and mutation processes), the data are constrained to a lower-dimensional manifold within this embedding space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets obey the manifold hypothesis.

@@

An estimate for $\rho_k$ is still needed. 
For this, we once again rely on the assumption that the query is drawn from the same distribution as the rest of the data, and thus the LFD at the query point is not significantly different from the LFD of nearby clusters.

@@

By Equation~\ref{eq:methods:lfd-original},

### 2024-01-04 sections/3-datasets.tex
tweaks from RFD

The benchmarks for Silva and RadioML were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM.
The OS kernel was Manjaro Linux 5.15.130-1-MANJARO.
The Rust compiler was Rust 1.72.0, and the Python interpreter version was 3.9.16.

\begin{table}[!t]
    % \renewcommand{\arraystretch}{1.15}
    \caption{Datasets used in benchmarks.}
    \label{tab:datasets:summary}
    \vskip 0.15in
    \begin{center}
        \begin{small}
            \begin{sc}
                \begin{tabular}{|l|l|l|l|}
                    \hline
                    \textbf{Dataset} & \textbf{Distance}  &\textbf{Cardinality}  & \textbf{Dimensionality}  \\
                    \hline
                    Fashion-Mnist    & Euclidean              & 60,000             & 784                    \\
                    \hline
                    Glove-25         & Cosine                 & 1,183,514          & 25                     \\
                    \hline
                    Sift             & Euclidean              & 1,000,000          & 128                    \\
                    \hline
                    Random           & Euclidean              & 1,000,000          & 128                    \\
                    \hline
                    Silva            & Levenshtein            & 2,224,640          & 3,712 - 50,000         \\
                    \hline
                    RadioML          & Dynamic Time Warping   & 97,920             & 1,024                  \\
                    \hline
                \end{tabular}
            \end{sc}
        \end{small}
    \end{center}
    \vskip -0.1in
\end{table}

@@

\subsection{Silva 18S}
\label{sec:datasets-and-benchmarks:silva-18s}

To showcase the use of CAKES with a more exotic distance function, we also benchmarked on the Silva 18S ribosomal DNA dataset~\cite{10.1093/nar/gks1219}.
This dataset contains ribosomal DNA sequences of 2,224,640 genomes with an aligned length of 50,000 letters.
We held out a set of 1,000 random sequences from the dataset to use as queries for benchmarking.
We use Levenshtein~\cite{levenshtein1966binary} distance on the unaligned sequences to build the tree and to perform $k$-NN search.
Since this dataset contains many \textit{redundant}, i.e. nearly identical, sequences, we used a maximum tree-depth of 128 as a partition \textit{criterion} in Algorithm~\ref{alg:methods:partition}.

>>>>

The benchmarks for SILVA and RadioML were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM.
The OS kernel was Manjaro Linux 5.15.130-1-MANJARO.
The Rust compiler was Rust 1.72.0, and the Python interpreter version was 3.9.16.

\begin{table}[!t]
    % \renewcommand{\arraystretch}{1.15}
    \caption{Datasets used in benchmarks.}
    \label{tab:datasets:summary}
    \vskip 0.15in
    \begin{center}
        \begin{small}
            \begin{sc}
                \begin{tabular}{|l|l|l|l|}
                    \hline
                    \textbf{Dataset} & \textbf{Distance}  &\textbf{Cardinality}  & \textbf{Dimensionality}  \\
                    \hline
                    Fashion-Mnist    & Euclidean              & 60,000             & 784                    \\
                    \hline
                    Glove-25         & Cosine                 & 1,183,514          & 25                     \\
                    \hline
                    Sift             & Euclidean              & 1,000,000          & 128                    \\
                    \hline
                    Random           & Euclidean              & 1,000,000          & 128                    \\
                    \hline
                    SILVA            & Levenshtein            & 2,224,640          & 3,712 - 50,000         \\
                    \hline
                    RadioML          & Dynamic Time Warping   & 97,920             & 1,024                  \\
                    \hline
                \end{tabular}
            \end{sc}
        \end{small}
    \end{center}
    \vskip -0.1in
\end{table}

@@

\subsection{SILVA 18S}
\label{sec:datasets-and-benchmarks:silva-18s}

To showcase the use of CAKES with a more exotic distance function, we also benchmarked on the SILVA 18S ribosomal DNA dataset~\cite{10.1093/nar/gks1219}.
This dataset contains ribosomal DNA sequences of 2,224,640 genomes with an aligned length of 50,000 letters.
We held out a set of 1,000 random sequences from the dataset to use as queries for benchmarking.
We use Levenshtein~\cite{levenshtein1966binary} distance on the unaligned sequences to build the tree and to perform $k$-NN search.
Since this dataset contains many \textit{redundant}, i.e. nearly identical, sequences, we used a maximum tree-depth of 128 as a partition \textit{criterion} in Algorithm~\ref{alg:methods:partition}.
